Particle Smoothing Algorithms for Variable Rate Models

### 2012-10-10 response_to_reviewers.tex

{\meta Ok. Now 100.}

- In Figs 7 and 10 the effective sample size would be a better measure of performance than the number of unique particles.

- The principal improvement offered by the more sophisticated particle smoothers compared to the filter-smoother is increased particle diversity. This should improve estimation performance but the improvements in consistency may be even greater. This is alluded to in the paper and partially covered by Figs 6 and 9 which show plots of sample trajectories. However, a more rigorous consistency analysis should be performed. For example, a simple approach would be to construct a Gaussian approximation to the posterior, establish a confidence ellipsoid and then count the number of times the true value falls within this ellipsoid.

{\meta These two comments concern the same question: How should particle approximations of a distribution be compared? There is, of course, much more to a distributional estimate than how close its mean is to the true value, so we do need something more than just the RMSE, as you suggest. One useful and established metric for measuring efficiency is the normalised estimation error squared (NEES), but we can't use this directly, because when the particle estimate is degenerate the NEES is infinite. Therefore, I've added a transformed version of the NEES as an efficiency measure. I think that together with the RMSE values, this provides sufficient evidence of the improvement provided by the smoother. }

@@

- Comparing the proposed method and the filter-smoother for equal sample sizes doesn't really make sense given that the former has $O(N^2)$ expense while the latter has $O(N)$ expense. The filter-smoother should be implemented with a much larger sample size such that both algorithms have approximately equal expense. This approach, which is taken in (Fearnhead et al, 2010), seems much fairer.

{\meta I think this is a bit misleading really. Although it might be a fairer test for a given problem, the result can be changed by modifying the scenario. For example, the relative performance of the smoother will be improved by having a longer data set. All I want to demonstrate in the simulations section is that the smoother operates as expected --- producing rejuvenated particle sets. The better algorithm for any application will depend on the model, the length of the data, and the desired output (e.g. is a distributional estimate needed, or is a single sample enough?).

In contrast to us, [Fearnhead et al. 2010] are comparing a variety of smoothing algorithms all of which do the same thing, produce a diverse set of marginal smoothing particles. Simple changes like altering the data length are less likely to affect their results, so their methods are appropriate to their purpose, but less so here.

However, Simon agrees with you. So i have actually changed the number of particles in the comparisons based on this equal-computation principle. }

### 2012-10-10 preprint_VRsmoothing.tex
Provisionally final second version

\documentclass[10pt,twocolumn,twoside]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

@@

\author{Pete~Bunch*,~\IEEEmembership{Member,~IEEE,}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

% The paper headers
\markboth{IEEE Transaction in Signal Processing}{IEEE Transaction in Signal Processing}%,~Vol.~1, No.~1, January~1901}%
%{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}

@@

\begin{IEEEkeywords}
Bayesian inference, state-space model, variable rate, particle filter, filtering, smoothing
\end{IEEEkeywords}

@@

\IEEEPARstart{T}{he} objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time. This is accomplished through the use of probabilistic models for the state evolution and measurement processes. Often, the latent state is a continuously varying quantity, whereas the observations are made at a discrete set of times. In these circumstances, it is simplest to discretise the state onto the same set of times as the observations. When the system is also Markovian, this leads to the standard ``fixed rate'' hidden Markov model (HMM). The standard HMM is poorly suited to systems where the state evolution contains discontinuities; for example, the price of a financial asset which may display large jumps at random times between periods of diffusion-like behaviour, or the kinematic state of a manoeuvring vehicle which may have sudden changes in the acceleration when turns begin or end. Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of changepoints which characterise transitions in behaviour.

@@

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the size of estimation errors of the resulting estimators. In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, a measure of estimator consistency is calculated by which to compare the algorithms, the mean empirical normalised estimation error squared (MENEES). This is defined as,  %the number of unique state samples is plotted over time in the following examples. A
%
\begin{IEEEeqnarray}{rCl}
 \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k (\hat{x}_k - x_k^*) \nonumber \\ 
 \hat{x}_k & = & \sum_i x_{k}^{(i)} \nonumber \\
 \hat{P}_k & = & \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
\end{IEEEeqnarray}

@@

\appendices
%\section{The Backwards Kalman Filter} \label{app:backwards_KF}
%
%A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian. Assume that at time $t_{n+1}$ there is a Gaussian density such that,
%%
%\begin{equation}
% p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
%\end{equation}
%
%The state transition model is linear-Gaussian, so the backwards predictive density is given by,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
%  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
%  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
%  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
%  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
%  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
%  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
%\end{IEEEeqnarray}
%
%where,
%%
%\begin{IEEEeqnarray}{rCl}
% \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
% \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
% \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
%\end{IEEEeqnarray}
%
%$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
%
%The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
% & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
% & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
% & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
%\end{IEEEeqnarray}
%
%where,
%%
%\begin{IEEEeqnarray}{rCl}
% \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
% \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
% \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
% \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
% \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
% \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
%\end{IEEEeqnarray}
%
%All steps use only standard Gaussian identities.

@@

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass{article}

\usepackage{IEEEtrantools}
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

@@

\author{Pete~Bunch and Simon~Godsill%
\thanks{Department of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}}

@@

@@

The objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time. This is accomplished through the use of probabilistic models for the state evolution and measurement processes. Often, the latent state is a continuously varying quantity, whereas the observations are made at a discrete set of times. In these circumstances, it is simplest to discretise the state onto the same set of times as the observations. When the system is also Markovian, this leads to the standard ``fixed rate'' hidden Markov model (HMM). The standard HMM is poorly suited to systems where the state evolution contains discontinuities; for example, the price of a financial asset which may display large jumps at random times between periods of diffusion-like behaviour, or the kinematic state of a manoeuvring vehicle which may have sudden changes in the acceleration when turns begin or end. Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of changepoints which characterise transitions in behaviour.

@@

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the size of estimation errors of the resulting estimators. In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, a measure of estimator consistency is calculated by which to compare the algorithms, the mean empirical normalised estimation error squared (MENEES). This is defined as,  %the number of unique state samples is plotted over time in the following examples. A
%
\begin{IEEEeqnarray}{rCl}
 \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k^{-1} (\hat{x}_k - x_k^*) \nonumber \\
 \hat{x}_k & = & \frac{1}{N_S} \sum_i x_{k}^{(i)} \nonumber \\
 \hat{P}_k & = & \frac{1}{N_S} \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
\end{IEEEeqnarray}

@@

\appendix
%\section{The Backwards Kalman Filter} \label{app:backwards_KF}
%
%A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian. Assume that at time $t_{n+1}$ there is a Gaussian density such that,
%%
%\begin{equation}
% p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
%\end{equation}
%
%The state transition model is linear-Gaussian, so the backwards predictive density is given by,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
%  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
%  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
%  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
%  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
%  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
%  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
%\end{IEEEeqnarray}
%
%where,
%%
%\begin{IEEEeqnarray}{rCl}
% \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
% \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
% \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
%\end{IEEEeqnarray}
%
%$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
%
%The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
% & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
% & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
% & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
%\end{IEEEeqnarray}
%
%where,
%%
%\begin{IEEEeqnarray}{rCl}
% \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
% \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
% \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
% \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
% \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
% \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
%\end{IEEEeqnarray}
%
%All steps use only standard Gaussian identities.

@@

\end{document}
